---
title: "Merging MOMS-PI dbGap Metadata"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Terminal Command: mount -t smbfs  //'rams;eid'@rams.adp.vcu.edu/som/shares/Biostatistics/Projects/MOMS-PI ~/local_folder_path/MOMS-PI


```{r cars}
rm(list=ls())
library(data.table)
library(tidyr)
#library(plyr)
library(dplyr)
library(stringr)
library(here)
library(phyloseq)
#library(HMP2Data) #install the library from github
```

## Including Plots

You can also embed plots, for example:

```{r files, echo=FALSE}

#path to protected dbGap data location

local_path <- "~/Documents/MOMS-PI"# change me
#local_path <- "~/MOMS-PI"# change me

path2 <- file.path(local_path,"71694", "PhenoGenotypeFiles", "RootStudyConsentSet_phs001523.MOMS_PI.v1.p1.c1.DS-PREG-COM-IRB-PUB-MDS", "PhenotypeFiles") 

load(here("data","stirrups_OTU.rda"))
load(here("data", "stirrups_samps.rda"))
physeq <- readRDS(here("data", "stirrups_phyloseq.RDS"))
newmoms <- rownames(meta_sub)
```

```{r Functions}
#Makes first row the column name and then deletes first row
#Input: Dataframe
#Output: Dataframe with one less row
header.true <- function(df) {
  names(df) <- as.character(unlist(df[1,]))
  df[-1,]
}
#Tests if any column is all NA
#Input: Dataframe
#Output: Nothing if no NA, otherwise outputs columns that are NA
na.test <-  function (x) {
  w <- sapply(x, function(x)all(is.na(x)))
  if (any(w)) {
    stop(paste("All NA in columns", paste(which(w), collapse=", ")))
  }
}
#Checks if df has any columns not in hope and updates to df to unique columns and a 'visit_ID' column
#Input: 2 Dataframes
#Output: Dataframe with unique columns
short.df <- function(df, hope) {
  ind <- which(names(df) %in% names(hope))
  visit <- which(names(df) == "visit_ID")
  count <- 0
  for (i in ind){
    i <- i - count
    if(i != (visit-count)){
      df <- df[,-i]
      count <- count + 1
    }
  }
  return(df)
}
```

On your secure server, create folder  MergedFiles inside phenotype data folder (where all phenotype data files are located) to save 16S HMP DACC data merged with dbGap phenotypes. 

```{r }
# Include specific file names that you want here:
# Add each file and merge to phyloseq object

#Read in clinical data file
clinical <- read.table(file.path(path2, "phs001523.v1.pht007510.v1.p1.c1.phenotype_clinical.DS-PREG-COM-IRB-PUB-MDS.txt"), fill = TRUE, header = FALSE, sep = "\t", dec = ".")
clinical <- header.true(clinical)#takes first row and makes it into colum names 
names(clinical)

#Read in health history file
health_history <- read.table(file.path(path2, "phs001523.v1.pht007512.v1.p1.c1.phenotype_healthHistorySurvey.DS-PREG-COM-IRB-PUB-MDS.txt"), fill = TRUE, header = FALSE, sep = "\t", dec = ".")
health_history <- header.true(health_history)#takes first row and makes it into colum names
#rename to have identifier for baseline health history collection visit
#rename(new variable name = existing variable name)
health_history <- health_history %>% rename("visit_num_health_survey" = "visit_num") 
#remove visit id at baseline to have history information appear for each visit
health_history <- health_history[,-(which(names(health_history) == "visit_ID"))] 
add_meta <- merge(x = clinical, y = health_history, all.x = TRUE)
#add_meta <- add_meta[order(add_meta$dbGaP_Subject_ID, add_meta$SUBJECT_ID, add_meta$visit_num),]

  
#In health history, some samples are labeled twice but have two different inputs. For instance, EP081373 has two different entries with one saying the subject is African American and the other one saying they are not.
discr = c("EP081373", "EP422409","EP936022", "EP989401")
discrepancies <- health_history[which(health_history$SUBJECT_ID %in% discr),]
#write.csv(discrepancies, file = file.path(path2, "MergedFiles", "Discrepancies.csv"))

na.test(add_meta) #Test if any column is all NA

#Read in review of systems file
revofsys <- read.table(file.path(path2, "phs001523.v1.pht007511.v1.p1.c1.phenotype_reviewOfSystemsSurvey.DS-PREG-COM-IRB-PUB-MDS.txt"), fill = TRUE, header = FALSE, sep = "\t", dec = ".")
revofsys <- header.true(revofsys)#takes first row and makes it into colum names 
revofsys <- short.df(revofsys, add_meta)
new_meta <- merge(x = add_meta, y = revofsys, all.x = TRUE)
na.test(new_meta)

#Read in sample descriptions file
multi <- read.table(file.path(path2, "phs001523.v1.pht007509.v1.p1.MOMS_PI_Sample.MULTI.txt"), fill = TRUE, header = FALSE, sep = "\t", dec = ".")
multi <- header.true(multi)#takes first row and makes it into colum names 
visit_id <- str_match(multi$SAMPLE_ID, "(.*_.*)_.*")[,2]
multi["visit_ID"] <- c(visit_id)
newmer <- merge(x = new_meta, y = multi, all = TRUE) 
length(which(!newmer$SAMPLE_ID %in% newmoms)) #8585 samples not in meta_sub file

#Merge to publicly available sample metadata
new_meta <- merge(x = meta_sub, y = newmer, by.x = 0, by.y = "SAMPLE_ID")
na.test(new_meta)


#Clean up data
new_data <- new_meta %>% distinct(Row.names, .keep_all = TRUE) #keeps only the first unique row name, done so that it can merge correctly with phyloseq object
new_data <- new_data[order(new_data$dbGaP_Subject_ID, new_data$visit_num),]
rownames(new_data) <- new_data$Row.names
new_data$Row.names <- NULL 

#race variable
race <- new_data[,28:33]
for (i in colnames(race)){
  race[[i]] <- recode(race[[i]], "Yes"= 1, "No"= 0)
}
race[] <- lapply(race, function(x) {
    if(is.factor(x)) as.numeric(as.character(x)) else x
})
race$sum <- rowSums(race, na.rm = TRUE)
race$race <- max.col(race[1:6])
race$race[which(race$sum == 0)] <- NA
race$race <- plyr::mapvalues(race$race, from = c(1:6), to = names(race[1:6]))
new_data$race <- race$race
#Merge to phyloseq object:
new_samps <- sample_data(new_data)

#construct the phyloseq 
stirrups_phyloseq_updated <- merge_phyloseq(physeq, new_samps)
#need to be saved to Biostats server
saveRDS(stirrups_phyloseq_updated, file = file.path(path2, "MergedFiles", "stirrups_phyloseq_updated.RDS"))


#Rows missing phenotypic information: 
rownames(new_data[is.na(new_data$african_american) & is.na(new_data$birthcontrol),]) #36 samples missing phenotypic information, 16 subjects in total
```


```{r }
filelist = list.files(path = path2, pattern = ".*.txt$")

init <- read.table(file.path(path2, filelist[9]), fill = TRUE, header = FALSE, sep = "", dec = ".")
init <- header.true(init)
names(init)

#See which from custom database are in the biostat server text file sample id's
check <- meta_sub[newmoms %in% init$SAMPLE_ID,] #2055 obs x 13 variables (so all samples in custom are captured)
samps <- init[init$SAMPLE_ID %in% newmoms,] #Create initial matrix for 2055 samples

#Read in second file
second <- read.table(file.path(path2, filelist[2]), fill = TRUE, header = FALSE, sep = "", dec = ".")
second <- header.true(second)
#Merge initial file and second file
comp_meta <- merge(samps, second)
#Create column for visit_ID as it is seen in most files and unique
visit_ID <- str_split_fixed(comp_meta$SAMPLE_ID, "_MV1D", n = 2)[,1]
comp_meta["visit_ID"] <- c(visit_ID)
#Read in first file
one <- read.table(file.path(path2, filelist[1]), fill = TRUE, header = FALSE, sep = "\t", dec = ".")
one <- header.true(one)
#Merge updated file and first file
comp_meta <- merge(comp_meta,one)
#For loop that goes through remaining file, besides last file, and merges the files based on the visit ID
for (i in 3:7) {
  file <- read.table(file.path(path2, filelist[i]), fill = TRUE, header = FALSE, sep = "\t", dec = ".")
  file <- header.true(file)
  file <- short.df(file, comp_meta) #Function to shorten file to just meaningful/unseen columns
  comp_meta <- merge(x = comp_meta, y = file, by = "visit_ID", all.x = TRUE)
}
#Merge last file
eight <- read.table(file.path(path2, filelist[8]), fill = TRUE, header = FALSE, sep = "\t", dec = ".")
eight <- header.true(eight)
comp_meta <- merge(x = comp_meta, y = eight, all.x = TRUE)
na.test(comp_meta) #Test if file has any columns with all NA values

#Merge into phyloseq
comp_samps <- sample_data(comp_meta)

stirrups_phyloseq_completed <- merge_phyloseq(physeq, comp_samps)

#have to be saved to Biostats server
saveRDS(stirrups_phyloseq_completed, file = file.path(path2, "MergedFiles", "stirrups_phyloseq_completed.RDS"))

```

To access slots of the phyloseq objects, use the following

```{r}
OTU_table <- as(otu_table(stirrups_phyloseq_updated), "matrix") 
Meta_data <- as(sample_data(stirrups_phyloseq_updated), "data.frame")
taxa_table <- as(tax_table(stirrups_phyloseq_updated), "matrix")

write.csv(OTU_table, file = file.path(path2, "MergedFiles", "OTU_table.csv"))
write.csv(Meta_data, file = file.path(path2, "MergedFiles", "Meta_data.csv"))


Meta_data_physeq <- as(sample_data(physeq), "data.frame")
write.csv(Meta_data_physeq, file = file.path(path2, "MergedFiles", "Meta_data_physeq.csv"))
physeq
```

