---
title: "Stirrups Phyloseq Creation"
output:
  html_document:
    df_print: paged
---

This file provides details to download and reshape the stirrupts data. 


## Stirrups data download workflow

* In order to download the stirrups data:
    + https://portal.hmpdacc.org, select "Data", Samples/Studies select MOMS-PI, Files/Format select "tbl", Files/Matrix Type select "16s_community"
    + Format: Study Name IS MOMS-PI AND File Format IS tbl AND File Matrix Type IS 16s_community
    + Download Stirrups data: ./hmp_client/bin/manifest2ascp.py --manifest=data.MOMS-PI/hmp_cart_2ae82fd042.tsv --user=username --password=password --ascp_path=/path/to/ascp/bin/ascp --ascp_options="-l 200M" > ascp-commands_biom_16S_MOMS-PI.sh

The shell script `ascp-commands_stirrups_MOMS-PI.sh` used  to download the stirrups data with aspera crient `Aspera cli` using terminal command line executable `ascp` is availale in the `Scripts/` folder. 

## Processing downloaded data

Each sample is downloaded as a separate file, which needs to be further reshaped into the single file (long format) that combines all samples using the set of unic commands documebted in `UnixScript.rtf` file in the `Scripts/` folder.

In terminal:

* merge all into one file using concatenate function
* Terminal Code: cat *.stirrups.profile.txt > MOMS_PI_allstirrups_quotes.txt"
* get rid of quotes, which cause issues later in R
* Note, you could use the -i flag to avoid another file, but this sometimes has issues on Macs
* Terminal Code: sed 's/\"//g' MOMS_PI_allstirrups_quotes.txt > MOMS_PI_allstirrups.txt"

```{r, echo=TRUE, results='hide'}
rm(list=ls())
library(data.table)
library(tidyr)
library(dplyr)
library(stringr)
library(phyloseq)
library(here)
```


## R Script to parse stirrups file

This script is used to parse the text file produced from the bash code into an OTU file of the stirrups file
This file is composed of 2055 samples of 360 taxa. This OTU matrix is then saved as an RDA file, which can be found on the github repository. 

```{r}
#Tests if any column is all NA
#Input: Dataframe
#Output: Nothing if no NA, otherwise outputs columns that are NA
na.test <-  function (x) {
  w <- sapply(x, function(x)all(is.na(x)))
  if (any(w)) {
    stop(paste("All NA in columns", paste(which(w), collapse=", ")))
  }
}

MOMSPI_stirrups<- as.data.frame(fread(here("data","MOMS_PI_allstirrups.txt")),sep="\t",header=FALSE,quote="")
colnames(MOMSPI_stirrups)<- c("Sample_Name","Taxon","Threshold_Status","No_Reads","Percent_Abundance","Average_Identity")
#For normalizing to only to those that had a hit (AT only)
#This might be the default, but we might also want to incorporate an option for all reads, including those without a hit
MOMSPI_stirrups_ATonly<-subset(MOMSPI_stirrups,MOMSPI_stirrups$`Threshold_Status`=="AT")
MOMSPI_stirrups_ATonly_forwide<-MOMSPI_stirrups_ATonly
MOMSPI_stirrups_ATonly_forwide$Threshold_Status<-NULL
MOMSPI_stirrups_ATonly_forwide$Percent_Abundance<-NULL
MOMSPI_stirrups_ATonly_forwide$Average_Identity<-NULL
momspi <- as.data.frame(spread(MOMSPI_stirrups_ATonly_forwide,key = Taxon, value = No_Reads))
rownames(momspi) <- momspi[,1]
momspi[,1] <- NULL
momspi[is.na(momspi)] <- 0
save(momspi, file = "stirrups_OTU.rda")
```

## Stirrups Taxa File Creation

This R scipt creates a taxa file composed only of the genus and species of the 360 taxa found in the OTU table. As information for the kingdom, phylum, class, order, and family for each taxa was unavailable, these columns were filled with NA. This script produces a matrix of 360 observations, each taxon, and 7 columns, taxonomic classification, which is then saved as an RDA file. This can be located on the github repository.

We will update this taxa table with all taxonomic information. 



```{r}
fName <- colnames(momspi)
both <- str_split_fixed(fName, "_", n = 2)
Genus <- both[,1]
Species <- both[,2]

Kingdom <- rep(NA, 360)
Phylum <- rep(NA ,360)
Class <- rep(NA, 360)
Order <- rep(NA, 360)
Family <- rep(NA, 360)
moms_taxa <- data.frame(Kingdom, Phylum, Class, Order, Family, Genus, Species, row.names = fName) #full taxa with genus, species, and full taxa name
taxa <- as.matrix(moms_taxa)
```

Save the final file to put on HMP2Data package

```{r}
save(taxa, file = "stirrups_taxa.rda")#make notes that this is the final file that goes into the HMP2Data package on github
```

## Stirrups Sample Metadata File Creation

This R scipt creates a metadata file composed of 13 columns of information (e.g visit_number, subject_race, subject_gender, etc.) for each of the 2055 samples. This script uses the information found in the momspi16S_samp.rda file to construct the metadata file for the 2055 Stirrups samples. 15 samples found in the Stirrups file were not listed in the momspi16S_samp file. Each of these samples were filled in with NA's for each column, as there was no further information on these samples. This leads to a matrix composed of 2055 observations of 13 variables. This is then saved into an RDA file, which can be located on the github repository.

Instructions on how to download rda using HMP2Data package. 

```{r}
#load("HMP2Data/data/momspi16S_mtx.rda")
load(here("data","momspi16S_samp.rda"))
#load("HMP2Data/data/momspi16S_tax.rda")

meta_sub <- momspi16S_samp[rownames(momspi16S_samp) %in% rownames(momspi),] 
#15 samples missing, fill in sample rows with NA's for these missing ones
missing <- which(!rownames(momspi) %in% rownames(meta_sub))
names <- c(rownames(momspi[missing,]))
meta_sub[nrow(meta_sub) + length(missing),] <- NA
firstNames <- rownames(meta_sub[1:2040,])
allNames <- c(firstNames, names)
rownames(meta_sub) <- allNames
#meta_sub should have all samples in the dataset
save(meta_sub, file = "stirrups_samps.rda")
```

## Stirrups Phyloseq Object Creation

This R script takes the Stirrups OTU, taxa, and sample metadata file and creates a phyloseq object from these three matrices. This is then saved as an RDS file, which is located on the github repository. 

```{r}
#Create phyloseq object:
OTU = otu_table(t(momspi), taxa_are_rows = TRUE)
TAX = tax_table(taxa)
SAMPLES = sample_data(meta_sub)
carbom <- phyloseq(OTU, TAX, SAMPLES)
saveRDS(carbom, file = "stirrups_phyloseq.RDS")
```
